Internal Gravity Wave Breaking in White Dwarf Binaries

### 2019-08-07 writeups/Su_IGW_break.tex
rename

We identify two instances of non-absorptive behavior: (i), the presence of a
reflected wave with wave vector $\bm{k} = k_{x}\bm{x} + k_{z}\bm{z}$, and
(ii) the amount of horizontal momentum flux $F$ reflected off the critical
layer.

@@

The measurements of $\hat{A}_i, \hat{A}_r, F_i, \Delta F, F_d, F_>$ are given in
\autoref{fig:nl_f_amps}. The oscillations in $\hat{A}_i, F_i, A_d, F_d$ are
apparent but the mean values seem to have temporally converged.

@@

Finally, we consider the impact of magnetic winding. In \citep{bukart}, magnetic
winding is used to enforce solid body rotation on the grounds that $t_A \gg
t_{gw}$, where
\begin{equation}
    t_A = \int\limits_0^R \frac{\sqrt{4\pi \rho}}{B_0}\;\mathrm{d}r
        \sim 10^2\;\mathrm{yr}\p*{\frac{10^3\;\mathrm{G}}{B}}.
\end{equation}
the Alfv\'en wave crossing time (evaluated for a CO WD in \citep{fullerIV})
measures the magnetic coupling time and $t_{gw}$ measures the gravitational wave
inspiral timescale. Before solid body rotation is attained, another relevant
timescale is the synchronization timescale $t_s$. For a tidal torque $\tau$ and
tidal forcing frequency $\sigma = m\p*{\Omega - \Omega_{spin}}$, we note that
angular momentum transfer is $\pd{M_{sync}}{t} \sigma R^2 = \tau$, where
$M_{sync}$ denotes the mass of the WD that has synchronized. Thus, the
synchronization timescale is
\begin{align}
    t_{sync} &\sim \frac{M_{sync}\sigma R^2}{\tau},\nonumber\\
        &\sim 2 \times 10^5\;\mathrm{yr}
            \p{\frac{M}{M_{\odot}}}
            \p{\frac{\sigma}{2\pi / (1\;\mathrm{hr})}}
            \p{\frac{R}{R_{\oplus}}}^2
            \p{\frac{10^{-14} GM_{\odot}^2/R_{\oplus}}{\tau}}.
\end{align}
A representative $\tau$ has been taken from \citep{bukart}. However, since only
the outer $\sim 10^{-4}M_{\odot}$ need be heated for the thermodynamically
interesting effects studied in \citep{fullerIV} and \citep{tidal_novae}, it
seems that magnetic winding cannot absolutely rule out energetic outbursts such
as tidal novae resulting from strong shear flows.

@@

\bibliographystyle{mnras}
\bibliography{paper}

@@

Numerically, the nonlinear $\frac{\bm{\nabla}P}{\rho}$ term is problematic: we
desire a system where the fluid fields are not divided by one another. We
introduce $\varpi = \frac{P}{\rho}$ instead, then mandate $\overline{\rho}, \varpi_0$
background fields satisfy hydrostatic equilibrium $\bm{\nabla}\varpi_0 +
\varpi_0 \bm{\nabla}\overline{\rho} + g\bm{z} = 0$. Taking isothermal stratification,
we find $\varpi_0 = gH$. We further change variables to $\Upsilon = \ln \rho -
\ln \overline{\rho}$ and $\varpi_1 = \varpi - \varpi_0$ deviations from the background
state to obtain a system of equations at most quadratic in fluid fields:
\begin{subequations}\label{se:nl_var}
    \begin{align}
        \bm{\nabla} \cdot \bm{u}_1 &= 0,\\
        \pd{\Upsilon}{t} + \p{\bm{u}_1 \cdot \bm{\nabla}} \Upsilon
            - \frac{u_z}{H} &= 0,\\
        \pd{u_{x}}{t} + \p{\bm{u}_1 \cdot \bm{\nabla}}u_{x}
            + \pd{\varpi_1}{x} + gH\pd{\Upsilon}{x}
            + \varpi_1 \pd{\Upsilon}{x} &= 0,\\
        \pd{u_z}{t} + \p{\bm{u}_1 \cdot \bm{\nabla}}u_z
            + \pd{\varpi_1}{z} + gH\pd{\Upsilon}{z}
            + \varpi_1 \pd{\Upsilon}{z} - \frac{\varpi_1}{H} &= 0.
    \end{align}
\end{subequations}
It bears noting that these equations are exactly equivalent to the original
Euler equations and hence conserve horizontal momentum.

@@

The nonlinear terms in the above equations will transfer energy from lower
wavenumbers to higher wavenumbers. Since spectral codes have no numerical
dissipation, artificial dissipation must be added. To ensure the dissipitive
system conserves horizontal momentum, we begin by adding dissipitive terms to
the flux-conservative form of the Euler fluid equations \autoref{se:nl_orig} (we
use stress tensor $\tau_{ij} = P\delta_{ij}$):
\begin{subequations}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\\
        \partial_t \rho + \bm{\nabla} \cdot (\rho \bm{u} - \nu
            \bm{\nabla}(\rho - \overline{\rho})) &= 0,\label{eq:visc_cons_mom}\\
        \partial_t (\rho \bm{u}) + \bm{\nabla} \cdot (\rho \bm{u}
        \bm{u} +
            \mathrm{diag}(\rho \varpi) - \nu \rho \bm{\nabla}\bm{u}) + \rho g
            \bm{z} &= 0.
    \end{align}
\end{subequations}
The same $\nu$ is used for both the diffusive and viscous term, though this is
not required. Since the dissipation is not physical and is purely used for
numerical stability, we choose it such that hydrostatic equilibrium is not
modified. Some algebraic manipulation to re-cast it in the form of
\autoref{se:nl_var} gives
\begin{subequations}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\\
        \partial_t \Upsilon + \p{\bm{u} \cdot \bm{\nabla}} \Upsilon -
            \frac{u_z}{H} - \nu\p{\nabla^2 \Upsilon + \p{\bm{\nabla}
            \Upsilon} \cdot \p{\bm{\nabla}\Upsilon} - \frac{2}{H}\partial_z
            \Upsilon + \frac{1 - e^{-\Upsilon}}{H^2}} &= 0,\\
        \partial_t \bm{u} + \p{\bm{u} \cdot \bm{\nabla}}\bm{u} +
            \bm{\nabla} \varpi + \varpi \bm{\nabla} \Upsilon - \nu \nabla^2
            \bm{u} + \bm{u} \nu\p{\nabla^2 \Upsilon + \p{\bm{\nabla}
            \Upsilon} \cdot \p{\bm{\nabla}\Upsilon} - \frac{2}{H}\partial_z
            \Upsilon + \frac{1 - e^{-\Upsilon}}{H^2}}&{}\nonumber\\
        - 2\nu \p{\p{\p{\bm{\nabla}\Upsilon} \cdot \bm{\nabla}}\bm{u} -
            \frac{1}{H}\partial_z \bm{u}} - \frac{\varpi_1}{H} &= 0.
    \end{align}
\end{subequations}
Hydrostatic equilibrium is still $\bm{\nabla} \varpi_0 + g\bm{z} = 0$ where
$\rho = \overline{\rho}, \bm{u} = 0$. Including the damping layers and forcing terms as
described in \autoref{ss:numerics}, we finally obtain the full system of
equations as simulated in Dedalus: [TODO update w/ mask]
\begin{subequations}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} ={}& 0,\\
        \partial_t \Upsilon - \frac{u_z}{H}
            ={}& \nu\p{\nabla^2 \Upsilon + \p{\bm{\nabla}
            \Upsilon} \cdot \p{\bm{\nabla}\Upsilon} - \frac{2}{H}\partial_z
            \Upsilon + \frac{1 - e^{-\Upsilon}}{H^2}},\nonumber\\
            & - \p{\bm{u} \cdot \bm{\nabla}}\Upsilon
                -\Gamma(z) \Upsilon
                + \frac{F}{\overline{\rho}(z)}e^{-\frac{(z - z_0)^2}{2\sigma^2}}
                    \cos \p{k_xx - \omega t},\\
        \pd{u_x}{t} + \pd{T}{x} + gH\pd{\Upsilon}{x} ={}&
            \nu \nabla^2 u_x
            - u_x \nu\p{\nabla^2 \Upsilon + \p{\bm{\nabla} \Upsilon} \cdot
                \p{\bm{\nabla}\Upsilon} - \frac{2}{H}\partial_z \Upsilon
                + \frac{1 - e^{-\Upsilon}}{H^2}}\nonumber\\
            &+ 2\nu \p{\p{\p{\bm{\nabla}\Upsilon} \cdot \bm{\nabla}}u_x
                - \frac{1}{H}\partial_z u_x}
            -\Gamma(z) u_x
                - \p{\bm{u} \cdot \bm{\nabla}}u_x
                - T_1 \pd{\Upsilon}{x},\\
        \pd{u_z}{t} + \pd{T}{z} + gH\pd{\Upsilon}{z} - \frac{T_1}{H} ={}&
            \nu \nabla^2 u_z
            - u_z \nu\p{\nabla^2 \Upsilon + \p{\bm{\nabla} \Upsilon} \cdot
                \p{\bm{\nabla}\Upsilon} - \frac{2}{H}\partial_z \Upsilon
                + \frac{1 - e^{-\Upsilon}}{H^2}}\nonumber\\
            &+ 2\nu \p{\p{\p{\bm{\nabla}\Upsilon} \cdot \bm{\nabla}}u_z -
                \frac{1}{H}\partial_z u_{z}}
            -\Gamma(z) u_z - \p{\bm{u} \cdot \bm{\nabla}}u_z
            - T_1 \pd{\Upsilon}{z}.
    \end{align}
\end{subequations}
\label{lastpage} % chktex 24
\end{document}

>>>>

We identify two manifestations of non-absorptive behavior: (i), the presence of
a reflected wave with wave vector $\bm{k} = k_{x}\bm{x} - k_{z}\bm{z}$, and (ii)
the amount of horizontal momentum flux $F$ reflected off the critical layer.
These two quantities need not agree exactly if some reflected flux is in
higher-order modes, which is observed. Both are of physical interest, however:
he reflected wave amplitude may be of physical interest in setting up standing
modes in a realistic star, while the reflected flux is important for accurately
tracking angular momentum transfer during synchronization.

@@

The measurements of $\hat{A}_i, \hat{A}_r, F_i, \Delta F, F_d, F_>$ are given in
\autoref{fig:nl_f_amps}. The oscillations in $\hat{A}_i, F_i, A_d, F_d$ are
significant but the mean values seem to have temporally converged.

@@

Finally, we consider the impact of magnetic winding. In \citep{bukart}, magnetic
winding is used to enforce solid body rotation on the grounds that $t_A \gg
t_{gw}$, where
\begin{equation}
    t_A = \int\limits_0^R \frac{\sqrt{4\pi \rho}}{B_0}\;\mathrm{d}r
        \sim 10^2\;\mathrm{yr}\p{\frac{10^3\;\mathrm{G}}{B}}.
\end{equation}
the Alfv\'en wave crossing time (evaluated for a CO WD in \citep{fullerIV})
measures the magnetic coupling time and $t_{gw}$ measures the gravitational wave
inspiral timescale. Before solid body rotation is attained, another relevant
timescale is the synchronization timescale $t_s$. For a tidal torque $\tau$ and
tidal forcing frequency $\sigma = m\p{\Omega - \Omega_{spin}}$, we note that
angular momentum transfer is $\pd{M_{sync}}{t} \sigma R^2 = \tau$, where
$M_{sync}$ denotes the mass of the WD that has synchronized. Thus, the
synchronization timescale is
\begin{align}
    t_{sync} &\sim \frac{M_{sync}\sigma R^2}{\tau},\nonumber\\
        &\sim 2 \times 10^5\;\mathrm{yr}
            \p{\frac{M}{M_{\odot}}}
            \p{\frac{\sigma}{2\pi / (1\;\mathrm{hr})}}
            \p{\frac{R}{R_{\oplus}}}^2
            \p{\frac{10^{-14} GM_{\odot}^2/R_{\oplus}}{\tau}}.
\end{align}
A representative $\tau$ has been taken from \citep{bukart}. However, since only
the outer $\sim 10^{-4}M_{\odot}$ need be heated for the thermodynamically
interesting effects studied in \citep{fullerIV} and \citep{tidal_novae}, it
seems that magnetic winding cannot absolutely rule out energetic outbursts such
as tidal novae resulting from strong shear flows.

@@

\bibliographystyle{mnras}
\bibliography{Su_IGW_break}

@@

Numerically, the nonlinear $\frac{\bm{\nabla}P}{\rho}$ term is problematic: we
desire a system where the fluid fields are not divided by one another. We
introduce $\varpi = \frac{P}{\rho}$ instead, then mandate $\overline{\rho},
\overline{\varpi}$ background fields satisfy hydrostatic equilibrium
$\bm{\nabla}\overline{\varpi} + \overline{\varpi} \bm{\nabla}\overline{\rho} +
g\bm{z} = 0$. Taking isothermal stratification, we find $\overline{\varpi} = gH$. We
further change variables to $\Upsilon = \ln \rho - \ln \overline{\rho}$ and
$\varpi' = \varpi - \overline{\varpi}$ deviations from the background state to obtain a
system of equations at most quadratic in fluid fields:
\begin{subequations}\label{se:nl_var}
    \begin{align}
        \bm{\nabla} \cdot \bm{u}' &= 0,\\
        \pd{\Upsilon}{t} + \p{\bm{u}' \cdot \bm{\nabla}} \Upsilon
            - \frac{u_z}{H} &= 0,\\
        \pd{u_{x}}{t} + \p{\bm{u}' \cdot \bm{\nabla}}u_{x}
            + \pd{\varpi'}{x} + gH\pd{\Upsilon}{x}
            + \varpi' \pd{\Upsilon}{x} &= 0,\\
        \pd{u_z}{t} + \p{\bm{u}' \cdot \bm{\nabla}}u_z
            + \pd{\varpi'}{z} + gH\pd{\Upsilon}{z}
            + \varpi' \pd{\Upsilon}{z} - \frac{\varpi'}{H} &= 0.
    \end{align}
\end{subequations}
It bears noting that these equations are exactly equivalent to the original
Euler equations and hence conserve horizontal momentum.

@@

The nonlinear terms in the above equations will transfer energy from lower
wavenumbers to higher wavenumbers. Since spectral codes have no numerical
dissipation, artificial dissipation must be added. To ensure the dissipitive
system conserves horizontal momentum exactly, we begin by adding dissipitive
terms to the flux-conservative form of the Euler fluid equations
\autoref{se:nl_orig} (we use stress tensor $\tau_{ij} = P\delta_{ij}$):
\begin{subequations}
    \begin{align}
        \bm{\nabla} \cdot \bm{u}' &= 0,\\
        \partial_t \rho + \bm{\nabla} \cdot (\rho \bm{u}' - \nu
            \bm{\nabla}(\rho - \overline{\rho})) &= 0,\label{eq:visc_cons_mom}\\
        \partial_t (\rho \bm{u}') + \bm{\nabla} \cdot (\rho \bm{u}' \bm{u}'
            + \mathrm{diag}(\rho \varpi)
            - \nu \rho \bm{\nabla}\bm{u})
            + \rho g \bm{z} &= 0.
    \end{align}
\end{subequations}
The same $\nu$ is used for both the diffusive and viscous term, though this is
not required. Since the dissipation is not physical and is purely used for
numerical stability, we choose it such that hydrostatic equilibrium is not
modified (hence $\nu$ acts only on $\rho - \overline{\rho}$).
% Some algebraic manipulation to re-cast it in the form of \autoref{se:nl_var}
% gives
% \begin{subequations}
%     \begin{align}
%         \bm{\nabla} \cdot \bm{u} &= 0,\\
%         \partial_t \Upsilon + \p{\bm{u} \cdot \bm{\nabla}} \Upsilon -
%             \frac{u_z}{H} - \nu\p{\nabla^2 \Upsilon + \p{\bm{\nabla}
%             \Upsilon} \cdot \p{\bm{\nabla}\Upsilon} - \frac{2}{H}\partial_z
%             \Upsilon + \frac{1 - e^{-\Upsilon}}{H^2}} &= 0,\\
%         \partial_t \bm{u} + \p{\bm{u} \cdot \bm{\nabla}}\bm{u} +
%             \bm{\nabla} \varpi + \varpi \bm{\nabla} \Upsilon - \nu \nabla^2
%             \bm{u} + \bm{u} \nu\p{\nabla^2 \Upsilon + \p{\bm{\nabla}
%             \Upsilon} \cdot \p{\bm{\nabla}\Upsilon} - \frac{2}{H}\partial_z
%             \Upsilon + \frac{1 - e^{-\Upsilon}}{H^2}}&{}\nonumber\\
%         - 2\nu \p{\p{\p{\bm{\nabla}\Upsilon} \cdot \bm{\nabla}}\bm{u} -
%             \frac{1}{H}\partial_z \bm{u}} - \frac{\varpi'}{H} &= 0.
%     \end{align}
% \end{subequations}
One last consideration we found necessary was masking out nonlinear terms in the
forcing zone with a similar form to \autoref{eq:Gamma}. In the absence of this
mask, a strong mean flow localized to the forcing zone developed. The mask used
was
\begin{equation}
    \Gamma_{NL}(z) = \frac{1}{2}\s{2
        + \tanh \frac{z - (z_0 + 8\sigma)}{\sigma}
        - \tanh \frac{z - z_B}{\sigma}}.
\end{equation}

Including the damping layers and forcing terms as described in
\autoref{ss:numerics}, we finally obtain the full system of equations as
simulated in Dedalus:
\begin{subequations}
    \begin{align}
        \bm{\nabla} \cdot \bm{u}' ={}& 0,\\
        \partial_t \Upsilon - \frac{u_z}{H}
            ={}& -\Gamma(z) \Upsilon
                + \frac{F}{\overline{\rho}(z)}e^{-\frac{(z - z_0)^2}{2\sigma^2}}
                    \cos \p{k_xx - \omega t},\nonumber\\
            & + \Gamma_{NL} \bigg[-\p{\bm{u}' \cdot \bm{\nabla}}\Upsilon
                + \nu\p{\nabla^2 \Upsilon + \p{
                    \bm{\nabla} \Upsilon} \cdot \p{\bm{\nabla}\Upsilon}
                    - \frac{2}{H}\partial_z \Upsilon
                    + \frac{1 - e^{-\Upsilon}}{H^2}}\bigg],\\
        \pd{u_x}{t} + \pd{\varpi'}{x} + gH\pd{\Upsilon}{x} ={}&
            -\Gamma(z) u_x
            + \Gamma_{NL}\bigg[\nu \nabla^2 u_x
            - u_x \nu\p{\nabla^2 \Upsilon + \p{\bm{\nabla} \Upsilon} \cdot
                \p{\bm{\nabla}\Upsilon} - \frac{2}{H}\partial_z \Upsilon
                + \frac{1 - e^{-\Upsilon}}{H^2}}\nonumber\\
            &+ 2\nu \p{\p{\p{\bm{\nabla}\Upsilon} \cdot \bm{\nabla}}u_x
                - \frac{1}{H}\partial_z u_x}
                - \p{\bm{u}' \cdot \bm{\nabla}}u_x
                - \varpi' \pd{\Upsilon}{x}\bigg],\\
        \pd{u_z}{t} + \pd{\varpi}{z} + gH\pd{\Upsilon}{z} - \frac{\varpi'}{H}
            ={}& -\Gamma(z) u_z
            +\Gamma_{NL}\bigg[\nu \nabla^2 u_z
            - u_z \nu\p{\nabla^2 \Upsilon + \p{\bm{\nabla} \Upsilon} \cdot
                \p{\bm{\nabla}\Upsilon} - \frac{2}{H}\partial_z \Upsilon
                + \frac{1 - e^{-\Upsilon}}{H^2}}\nonumber\\
            &+ 2\nu \p{\p{\p{\bm{\nabla}\Upsilon} \cdot \bm{\nabla}}u_z -
                \frac{1}{H}\partial_z u_{z}}
            - \p{\bm{u}' \cdot \bm{\nabla}}u_z
            - \varpi' \pd{\Upsilon}{z}\bigg].
    \end{align}
\end{subequations}
\label{lastpage} % chktex 24
\end{document}
